Safe Transferable Regions

### 2016-10-07 paper/type-inference.tex
type inference updates

\newcommand{\soln}{\eta}
\newcommand{\saturate}[1]{{#1}^*}
\newcommand{\groundc}[1]{\saturate{#1}_g}
\newcommand{\consOf}[1]{\textsc{GenConstraint}(#1)}
\newcommand{\satC}{\saturate{C}}

@@

\paragraph{Constraint Solver.}
Our algorithm for solving a set of constraints $C$ consists of the following steps:
\begin{enumerate}
\item We use an iterative fixed point (saturation) algorithm to compute
$\saturate{C}$, the set of all constraints logically implied by $C$.
\item Let $\groundc{C}$ denote the subset of all ground constraints
(\ie, constraints without any region variable or predicate variable)
in $\saturate{C}$. We check if $\groundc{C}$ is valid in the theory of
partial-order. If $\groundc{C}$ is valid, the solution $\soln$ is determined
as below. Otherwise, the constraint solver fails, as no valid solution exists.
\begin{itemize}
\item For any predicate variable $\varphi$, $\predSubstFn(\varphi)$ is defined
to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
\item For any region variable $\rho$, we define $\regionSubstFn(\rho)$ to be any element of the
set $\{ \pi \in \regionDeltaMap(\rho) \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$ 
if this set is non-empty.
Otherwise, the solver fails.
%\item For any region variable $\rho$,  we are guaranteed that there exists a unique
% $\pi \in \regionConstants$  such that $\saturate{C}$ contains some constraint
%with a consequent $\rho = \pi$. We define $\regionSubstFn(\rho)$ to be $\pi$.
\end{itemize}
\end{enumerate}

@@

Correctness follows by establishing:
\begin{theorem}
(a) Any assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
(b) $\satC$ is satisfiable iff $\groundc{C}$ is valid.
(c) For any region variable $\rho$, $\satC$ contains some constraint of the
form $\isvalid{\ell}{\rho = \pi}$.
\end{theorem}

The last item in the above theorem follows because every region variable is guaranteed to be unified with some
region constant by the generated constraints\footnote{
There is one special case where this does not hold: a recursive function
that calls itself in a non-terminating fashion. Such a function never
returns a value, and so the return-value can be typed as anything.
We assume that the return value of such a function is typed to be \C{unit},
which will not introduce any region variables at the call-site of such a function.}
This follows because $\FB$ does not admit uninitialized variables.
% , and the type system uniquely determines the region parameters of every value

@@

>>>>

\newcommand{\soln}{\eta}
\newcommand{\saturate}[1]{{#1}^*}
\newcommand{\myground}[1]{\saturate{#1}_g}
\newcommand{\consOf}[1]{\textsc{GenConstraint}(#1)}
\newcommand{\satC}{\saturate{C}}
\newcommand{\groundC}{\myground{C}}

@@

\paragraph{Constraint Solver.}
Our algorithm for solving a set of constraints $C$ consists of the following steps:
\begin{enumerate}
\item We use an iterative fixed point (saturation) algorithm to compute
$\saturate{C}$, the set of all constraints logically implied by $C$.
\item Let $\myground{C}$ denote the subset of all ground constraints
(\ie, constraints without any region variable or predicate variable)
in $\saturate{C}$. We check if $\myground{C}$ is valid in the theory of
partial-order. If $\myground{C}$ is valid, the solution $\soln$ is determined
as below. Otherwise, the constraint solver fails, as no valid solution exists.
\begin{itemize}
\item For any predicate variable $\varphi$, $\predSubstFn(\varphi)$ is defined
to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
\item For any region variable $\rho$, we define $\regionSubstFn(\rho)$ to be any element of the
set $\{ \pi \in \regionDeltaMap(\rho) \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$ 
if this set is non-empty.
Otherwise, the solver fails.
%\item For any region variable $\rho$,  we are guaranteed that there exists a unique
% $\pi \in \regionConstants$  such that $\saturate{C}$ contains some constraint
%with a consequent $\rho = \pi$. We define $\regionSubstFn(\rho)$ to be $\pi$.
\end{itemize}
\end{enumerate}

@@

% Correctness follows by establishing:
\begin{theorem}
  \label{thm:closure}
Let $C = \consOf{q}$.
Any assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
%% (b) $\satC$ is satisfiable iff $\myground{C}$ is valid.
%% (c) For any region variable $\rho$, $\satC$ contains some constraint of the
%% form $\isvalid{\ell}{\rho = \pi}$.
\end{theorem}

\begin{proof}
  For the forward implication, we use induction to show that every constraint
  added to $\satC$ preserves satisfiability (with respect to an assignment $\soln$).
  The Transitivity and Substitution steps are straightforward. The Abduction Decomposition
  step is justified by Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
  The reverse implication is trivial.
\end{proof}

@@

\begin{proof}

Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.

\TODO{The following is an informal sketch, requiring some formalization.}

The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
It is easy to show that we take care of these.

The more difficult case are validity constraints, involving either region variables
($\rho$) or predicate variables ($\varphi$) or both.

The constraint-generation rules guarantee that every region variable will
be unified with some region constant\footnote{
There is one special case where this does not hold: a recursive function
that calls itself in a non-terminating fashion. Such a function never
returns a value, and so the return-value can be typed as anything.
We assume that the return value of such a function is typed to be \C{unit},
which will not introduce any region variables at the call-site of such a function.}
This follows because $\FB$ does not admit uninitialized variables.
% , and the type system uniquely determines the region parameters of every value

We use this property to show that it is okay to ignore validity constraints
in $\satC$ containing region variables. For every
region variable $\rho$ unified with a constant $\pi$, and for every validity
constraint containing $\rho$, we will have a corresponding validity constraint
containing $\pi$ instead.
Note: We likely need to add another rule to the saturation rules for computing
$\satC$, namely $\isvalid{\ell}{r} \in \satC \Rightarrow \isvalid{\ell \conj \phi}{r} \in \satC$.
We need this to propagate constraints from an outer scope into an inner scope.

As for validity constraints containing a predicate variable $\varphi$ on the consequent,
we argue that the use of the Substitution rule in computing $\satC$ makes it okay
to ignore these constraints.

As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
us convert these into simpler constraints that do not have $\varphi$ on the
antecedent or into a constraint that is directly incorporated into definition
of $\predSubstFn$.
\end{proof}
